Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-14 analysis/paper/paper.Rmd
updates for background and discussion

The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs. The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. Two measurements for thickness and diameter for each pot were took to obtain the average value. To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. The results show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases. The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%). For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period(10.6%).

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [\citation].

```{r tidy-data}
# Once we have the jpgs, start from here
library(tidyverse)
library(here)

@@

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by the metric attributes, specifically the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the only attribute that presents homogeneous. In addition to the emergence of specialized groups indicative of changes in social organization, another explanation is that producing homogeneous pottery is an act of resistance to the Chinese immigrant by emphasizing the practice of traditional pottery production. Among the pottery attributes, shape is the most  obvious and visible trait, which can be used as an expression of social boundary or group identity [cf. @Roux2015].

>>>>

```{r metric-CVs-test}
library(here)
metric_cvs_with_phase <- readr::read_csv(here("analysis", "data", "raw_data", "Kwl_p_metrics_cv_test.csv"))

knitr::kable(metric_cvs_with_phase,
             caption = "Coefficient of Variation test for metric attributes by phases")
```

The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs. The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. The results (Table \@ref(tab:metric-CVs-test))show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases. The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%). For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period(10.6%).

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. It seems there is no obvious changes in the inclusions over time.

```{r tidy-data}
# Once we have the jpgs, start from here
library(tidyverse)

@@

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by most of the metric attributes, such as the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the main attribute that presents homogeneity. In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the the Chinese immigrants by emphasizing the practice of traditional pottery production. In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneous and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southeast China into Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identity, especially in the context of Han immigrants [@Wu2019]. At Kiwulan, among the pottery attributes, shape is the most obvious and visible trait, which could be viewed as an expression of social boundary or identity [cf. @Roux2015].
